Block bootstrap consistency\\under weak assumptions

### 2016-09-14 Paper.tex
[#28] Tweak intro to proof of theorem 1

\subsection*{Proof of Theorem~\ref{T1}}
{%
\newcommand{\Db}[1][j]{D_{n#1}^{*}}
\newcommand{\Zsum}{\sum_{j=1}^{J_n} \Zb / \sigma_n}
Observe that
\[
\sqrt{n}(\bar X_{n}^{*} -  \E^{*} \bar X_n^{*})
= \sqrt{n}((\bar{X}_{n}^{*} - \bar{\mu}_{n}^{*}) - \E^{*} (\bar X_n^{*} - \bar{\mu}_n^{*}))
+ \sqrt{n}(\bar{\mu}_{n}^{*} - \E^{*} \bar{\mu}_n^{*}).
\]
Lemma~\ref{L3} implies that the last term is $o_{L_2}(1)$, so we can assume that
$\mu_{nt}$ is constant in $t$ for each $n$ for the proof of normality. Convergence
in probability of the bootstrap variance and the bootstrap variance estimator
holds by Lemma~\ref{L4}.

@@

To prove~\eqref{eq:19}, it suffices to show that
\begin{equation}\label{eq:21}
  \Em \sum_{j=1}^{J_n} \Em^{*}\big(Z_{nj}^{*2} \ind\{Z_{nj}^{*2} > \epsilon\}\big) \to^p 0.
\end{equation}
For any $j$ and $m$ and for large enough $n$,
\begin{align}
\Em \Em^{*}\big((Z_n(u_{nj}, m)^2n/m) &\ind\{Z_n(u_{nj}, m)^2n/m > \epsilon n/m\}\big) \notag \\
&= \E \big((Z_n^{*}(0, m)^2n/m) \ind\{Z_n^{*}(0, m)^2n/m > \epsilon n/m\}\big) \notag \\
&\leq B(\epsilon n/m)\label{eq:5}
\end{align}
where $B$ is some finite monotone function such that $B(x) \to 0$ as
$x \to \infty$; the existence of a function $B$ that satisfies these
conditions (and does not depend on $n$ or $m$) is a consequence of
Lemma~\ref{L6},
Equation~\eqref{eq:30}. Since $J_n$ and $M_{nj}$ are both $\Ms$-measurable
and independent of the $u_{nj}$'s and $X_{nt}$'s,~\eqref{eq:5} implies that
\begin{align*}
\Em &\sum_{j=1}^{J_n} \Em^{*}\big(Z_{nj}^{*2} \ind\{Z_{nj}^{*2} > \epsilon\}\big) \\
& = \sum_{j=1}^{J_n} (M_{nj}/n) \Em \Em^{*}
\big((Z_n(u_{nj}, M_{nj})^2 n/M_{nj})
\ind\{Z_n(u_{nj}, M_{nj})^2n/M_{nj} > \epsilon n/M_{nj}\}\big) \\
& \leq \sum_{j=1}^{J_n} (M_{nj}/n) \, B(\epsilon n/M_{nj}) \\
& \leq \max_{j = 1,\dots,J_n} B(\epsilon n / M_{nj}) \sum_{i=1}^{J_n} M_{ni} / n \\
& = B(\epsilon n / \max_{j = 1,\dots,J_n} M_{nj}) \\
& \to^p 0 \text{\ as\ } n \to \infty.
\end{align*}
The first equality follows from simple algebra and the definition
of the $u_{nj}$'s. The first inequality is a
consequence of~\eqref{eq:5}.  The second equality holds by
monotonicity of $B$. And convergence in probability holds by
Lemma~\ref{L1}, completing the proof
of~\eqref{eq:19}. The Dominated Convergence Theorem and~(\ref{eq:18})
then ensure that
\begin{equation}\label{eq:22}
  \pr^{*} \Big[\sum_{j=1}^{J_n}
  Z_{nj}^{*} / \sigma_n^* \leq x \Big] - \Phi(x) \to^p 0.
\end{equation}
(Also see
Lemma~\ref{L2}.)

@@

  For part 5, write
  \begin{equation}\label{eq:41}
    \sum_{i=1}^{J_n} M_{ni}^2 =
    \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})^2
    + \tfrac{2}{p_n} \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})
    + \tfrac{J_n}{p_n^2}
  \end{equation}
  so
  \begin{equation*}
    \E \Big( \sum_{i=1}^{J_n} M_{ni}^2 \Big)
    = \E\Big( \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})^2 \Big)
    + \tfrac{2 n}{p_n} - \E \tfrac{J_n}{p_n^2}.
  \end{equation*}
  Since $J_n$ is a stopping time and $\sum_{i=1}^j (M_{ni} - 1/p_n)$
  is a martingale, we have the equality \citep[Lemma 6]{CRT:65}
  \begin{align*}
    \E\Big( \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})^2 \Big)
    &= \E\Big( \sum_{i=1}^{J_n} \E((M_{ni} - \tfrac{1}{p_n})^2 \mid M_{n1},\dots,M_{n,i-1})\Big)\\
    &= \tfrac{1 - p_n}{p_n^2} \E J_n.
  \end{align*}
  Then $\E J_n = (n-1) p_n + 1$, completing the proof.
\end{proof}

@@

\begin{lem}\label{L3}
Under the conditions of Theorem~\ref{T1},
\begin{equation}\label{eq:40}
  \E \Big(\sum_{j=1}^{J_n} \Big( \sum_{t = K_{n,j-1}+1}^{K_{nj}}
  (\mu_{nt}^{*} - \bar \mu_n)\Big)^2\Big) = o(n p_n)
\end{equation}
and $\bar \mu_n^{*} = \E^{*} \bar \mu_n^{*} + o_{L_2}(n^{-1/2})$.
\end{lem}
\begin{proof}
  Since the blocks are independent and
  $\E^{*} \bar\mu_n^{*} = \bar\mu_n$ under the stationary bootstrap,
  the result on $\bar\mu_n^{*}$ follows from~\eqref{eq:40}. To
  show~\eqref{eq:40}, observe that
  \begin{align*}
    \E \Big(\sum_{j=1}^{J_n} &\Big( \sum_{t= K_{n,j-1} +1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar \mu_n)\Big)^2\Big)= \\
    &= \E\Big(\sum_{j=1}^{J_n}\sum_{s,t = K_{n,j-1}+1}^{K_{nj}}
    \E \big((\mu_{ns}^{*} - \bar{\mu}_n)
    (\mu_{nt}^{*} - \bar{\mu}_n) \mid \Ms \big)\Big)\\
    &= \E\Big( \sum_{j=1}^{J_n}
    \sum_{k,l = -n}^{n} \tfrac{1}{n} \sum_{\tau=0}^{n-1}
    (\mu_{n,\tau+k} - \bar{\mu}_n) (\mu_{n,\tau+l} - \bar{\mu}_n) \;
    1\{ \tau + k, \tau + l \in I_n(\tau, M_{nj})\}\Big)\\
    &\leq \E\Big(\sum_{j=1}^{J_n}
    \Big(\sum_{k = -n}^{n} \Big(\tfrac{1}{n} \sum_{\tau=0}^{n-1}
    (\mu_{n,\tau+k} - \bar{\mu}_n)^2
    1\{ \tau + k \in I_n(\tau, M_{nj})\}\Big)^{1/2}\Big)^{2}\Big)\\
    &= \E\Big(\sum_{j=1}^{J_n}
    M_{nj}^2 \Big\lvert \tfrac{1}{n} \sum_{\tau=0}^{n-1}
    (\mu_{n,\tau} - \bar{\mu}_n)^2 \Big\lvert\Big)\\
    &= o(n p_n).
  \end{align*}
  The first line follows from the LIE. The second exploits the
  fact that the conditional expectation in the previous line
  just averages over the block's starting period. The
  inequality is an application of the Cauchy-Schwarz inequality.
  The next line holds because the summation
  \[
    \sum_{\tau=0}^{n-1} (\mu_{n,\tau+k} - \bar{\mu}_n)^2
    1\{ \tau + k \in I_n(\tau, M_{nj})\}
  \]
  is zero for all but $M_{nj}$ values of $k$ and is a nonzero constant
  for those $k$. The last line holds by Lemma~\ref{L1} and assumption.
\end{proof}

@@

\begin{proof}
  For~\eqref{eq:25}, $(Z_{nj}^{*2} - \Em^* Z_{nj}^{*2}) \cdot (n /
  M_{nj})$ is a uniformly integrable martingale difference array, by
  Lemma~\ref{L6}, and
  satisfies the \lln. (See~\citealp{Dav:94}, Theorem 19.7.)
  For~\eqref{eq:26}, observe that
  \begin{multline*}
    \Big| \sum_{j=1}^{J_n} (Z_{nj}^{*2} - Z_{nj}^{\prime*2}) \Big|
    \leq 2 \Big(\sum_{j=1}^{J_n} Z_{nj}^{*2} \Big)^{1/2}
    \Big(\tfrac{1}{n} \sum_{j=1}^{J_n} \Big( \sum_{t= K_{n,j-1} + 1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar X_n)\Big)^2 \Big)^{1/2} \\
    + \tfrac{1}{n} \sum_{j=1}^{J_n} \Big( \sum_{t= K_{n,j-1} + 1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar X_n)\Big)^2
  \end{multline*}
  from the Cauchy-Schwarz inequality. Lemma~\ref{L5}, along
  with~\eqref{eq:17} and~\eqref{eq:25}, implies that
  $\sum_j Z_{nj}^{*2} = O_p(1)$. Lemma~\ref{L3} implies that
  \[
    \tfrac{1}{n} \sum_{j=1}^{J_n} \Big( \sum_{t= K_{n,j-1} + 1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar X_n)\Big)^2 \to^p 0
  \]
  since $\bar X$ itself obeys the \lln.

@@

\begin{proof}
  We will first show that, for every $\epsilon > 0$ and large enough
  $n$, there exists an $\epsilon' > 0$ with the property that
  \begin{equation}\label{eq:31}
    \E \Big(\ind(A) \times \max_{m = 1,\dots,m'} Z^*_n(\tau, m)^2 n / m'\Big) < \epsilon
  \end{equation}
  for any event $A \in \Fs_n \equiv \sigma(X_{n1},\dots,X_{nn})$ with
  $\pr(A) \leq \epsilon'$. Then we will show that this property
  implies uniform integrability.\footnote{%
    The novelty in this part of the proof is the measurability
    requirement. The sigma-field $\Fs_n$ intentionally excludes the
    random variables that define the bootstrap (block length and block
    start periods) and any $A \in \Fs_n$ is independent of those
    variables, so $Z_n^{*}(\tau, m)^2$ is not $\Fs_n$-measurable. If
    $\Fs_n$ were enlarged to include those variables then uniform
    integrability would be an immediate consequence and would not need
    to be shown separately. See \citet[Theorem 12.9]{Dav:94}.} %

>>>>

\subsection*{Proof of Theorem~\ref{T1}}
{%
\newcommand{\Db}[1][j]{D_{n#1}^{*}}
\newcommand{\Zsum}{\sum_{j=1}^{J_n} \Zb / \sigma_n}

@@

To prove~\eqref{eq:19}, it suffices to show that
\begin{equation}\label{eq:21}
  \Em \sum_{j=1}^{J_n} \Em^{*}\big(Z_{nj}^{*2} \ind\{Z_{nj}^{*2} > \epsilon\}\big) \to^p 0.
\end{equation}
For any $j$ and $m$ and for large enough $n$,
\begin{align}
\Em \Em^{*}\big((Z_n(u_{nj}, m)^2n/m) &\ind\{Z_n(u_{nj}, m)^2n/m > \epsilon n/m\}\big) \notag \\
&= \E \big((Z_n^{*}(0, m)^2n/m) \ind\{Z_n^{*}(0, m)^2n/m > \epsilon n/m\}\big) \notag \\
&\leq B(\epsilon n/m)\label{eq:5}
\end{align}
where $B$ is some finite monotone function such that $B(x) \to 0$ as
$x \to \infty$; the existence of a function $B$ that satisfies these
conditions (and does not depend on $n$ or $m$) is a consequence of
Lemma~\ref{L6}. Since $J_n$ and $M_{nj}$ are both $\Ms$-measurable
and independent of the $u_{nj}$'s and $X_{nt}$'s,~\eqref{eq:5} implies that
\begin{align*}
\Em &\sum_{j=1}^{J_n} \Em^{*}\big(Z_{nj}^{*2} \ind\{Z_{nj}^{*2} > \epsilon\}\big) \\
& = \sum_{j=1}^{J_n} (M_{nj}/n) \Em \Em^{*}
\big((Z_n(u_{nj}, M_{nj})^2 n/M_{nj})
\ind\{Z_n(u_{nj}, M_{nj})^2n/M_{nj} > \epsilon n/M_{nj}\}\big) \\
& \leq \sum_{j=1}^{J_n} (M_{nj}/n) \, B(\epsilon n/M_{nj}) \\
& \leq \max_{j = 1,\dots,J_n} B(\epsilon n / M_{nj}) \sum_{i=1}^{J_n} M_{ni} / n \\
& = B(\epsilon n / \max_{j = 1,\dots,J_n} M_{nj}) \\
& \to^p 0 \text{\ as\ } n \to \infty.
\end{align*}
The first equality follows from simple algebra and the definition
of the $u_{nj}$'s. The first inequality is a
consequence of~\eqref{eq:5}.  The second equality holds by
monotonicity of $B$. And convergence in probability holds by
Lemma~\ref{L1}, completing the proof
of~\eqref{eq:19}. The Dominated Convergence Theorem and~(\ref{eq:18})
then ensure that
\begin{equation}\label{eq:22}
  \pr^{*} \Big[\sum_{j=1}^{J_n}
  Z_{nj}^{*} / \sigma_n^* \leq x \Big] - \Phi(x) \to^p 0.
\end{equation}
(Also see
Lemma~\ref{L2}.)

@@

  For part 5, write
  \begin{equation}\label{eq:41}
    \sum_{i=1}^{J_n} M_{ni}^2 =
    \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})^2
    + \tfrac{2}{p_n} \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})
    + \tfrac{J_n}{p_n^2}
  \end{equation}
  so
  \begin{equation*}
    \E \Big( \sum_{i=1}^{J_n} M_{ni}^2 \Big)
    = \E\Big( \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})^2 \Big)
    + \tfrac{2 n}{p_n} - \E \tfrac{J_n}{p_n^2}.
  \end{equation*}
  Since $J_n$ is a stopping time and $\sum_{i=1}^j (M_{ni} - 1/p_n)$
  is a martingale, we have the equality \citep[Lemma 6]{CRT:65}
  \begin{align*}
    \E\Big( \sum_{i=1}^{J_n} (M_{ni} - \tfrac{1}{p_n})^2 \Big)
    &= \E\Big( \sum_{i=1}^{J_n} \E((M_{ni} - \tfrac{1}{p_n})^2 \mid M_{n1},\dots,M_{n,i-1})\Big)\\
    &= \tfrac{1 - p_n}{p_n^2} \E J_n.
  \end{align*}
  Finally, $\E J_n = (n-1) p_n + 1$, completing the proof.
\end{proof}

@@

\begin{lem}\label{L3}
Under the conditions of Theorem~\ref{T1},
\begin{equation}\label{eq:40}
  \E \Big(\sum_{j=1}^{J_n} \Big( \sum_{t = K_{n,j-1}+1}^{K_{nj}}
  (\mu_{nt}^{*} - \bar \mu_n)\Big)^2\Big) = o(n p_n)
\end{equation}
and $\bar \mu_n^{*} = \E^{*} \bar \mu_n^{*} + o_{L_2}(n^{-1/2})$.
\end{lem}
\begin{proof}
  The second result, on $\bar\mu_n^*$ is an immediate implication of
  Equation~\eqref{eq:40}. To show~\eqref{eq:40}, observe that
  \begin{align*}
    \E \Big(\sum_{j=1}^{J_n} &\Big( \sum_{t= K_{n,j-1} +1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar \mu_n)\Big)^2\Big)= \\
    &= \E\Big(\sum_{j=1}^{J_n} \big\lvert \sum_{s,t = K_{n,j-1}+1}^{K_{nj}}
    \E \big((\mu_{ns}^{*} - \bar{\mu}_n)
    (\mu_{nt}^{*} - \bar{\mu}_n) \mid \Ms \big) \big\rvert \Big)\\
    &\leq \E\Big( \sum_{j=1}^{J_n}
    \sum_{k,l = -n}^{n} \tfrac{1}{n} \sum_{\tau=0}^{n-1} \big\lvert
    (\mu_{n,\tau+k} - \bar{\mu}_n) (\mu_{n,\tau+l} - \bar{\mu}_n) \;
    1\{ \tau + k, \tau + l \in I_n(\tau, M_{nj})\} \big\rvert \Big)\\
    &\leq \E\Big(\sum_{j=1}^{J_n} \Big[
    \sum_{k = -n}^{n} \big(\tfrac{1}{n} \sum_{\tau=0}^{n-1}
    (\mu_{n,\tau+k} - \bar{\mu}_n)^2
    1\{ \tau + k \in I_n(\tau, M_{nj})\}\big)^{1/2}\Big]^{2}\Big)\\
    &= \E\Big(\sum_{j=1}^{J_n}
    M_{nj}^2 \Big\lvert \tfrac{1}{n} \sum_{\tau=0}^{n-1}
    (\mu_{n,\tau} - \bar{\mu}_n)^2 \Big\lvert\Big)\\
    &= o(n p_n).
  \end{align*}
  The first equality follows from the LIE. The first inequality
  exploits the fact that the conditional expectation in the previous
  expression averages over the block's starting period but treats
  every other random variable as contstant. (The argument of the
  indicator function and the indices of the summations are
  deliberately chosen to accomodate blocks that include observations
  from both the beginning and the end of the series.) The second
  inequality is an application of the Cauchy-Schwarz inequality to the
  inner-most summation. The next equality holds because the
  summation
  \[
    \sum_{\tau=0}^{n-1} (\mu_{n,\tau+k} - \bar{\mu}_n)^2
    1\{ \tau + k \in I_n(\tau, M_{nj})\}
  \]
  is zero for all but $M_{nj}$ values of $k$ and is a nonzero constant
  for those $k$. The last line holds by Lemma~\ref{L1} and assumption.
\end{proof}

@@

\begin{proof}
  For~\eqref{eq:25}, $(Z_{nj}^{*2} - \Em^* Z_{nj}^{*2}) \cdot (n /
  M_{nj})$ is a uniformly integrable martingale difference array, by
  Lemma~\ref{L6}, and
  satisfies the \lln. (See~\citealp{Dav:94}, Theorem 19.7.)
  For~\eqref{eq:26}, observe that
  \begin{align*}
    \Big| \sum_{j=1}^{J_n} (Z_{nj}^{*2} - Z_{nj}^{\prime*2}) \Big|
    &= \Big| \sum_{j=1}^{J_n} \big(Z_{nj}^{*} - \big[Z_{nj}^{*} + \tfrac{1}{\sqrt{n}} \sum_{t=K_{n,j-1}+1}^{K_{nj}} (\bar X_n - \mu_{nt}^*)\big]^2\big) \Big| \\
    &\leq 2 \Big(\sum_{j=1}^{J_n} Z_{nj}^{*2} \Big)^{1/2}
    \Big(\tfrac{1}{n} \sum_{j=1}^{J_n} \Big( \sum_{t= K_{n,j-1} + 1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar X_n)\Big)^2 \Big)^{1/2}\\
    &\quad + \tfrac{1}{n} \sum_{j=1}^{J_n} \Big( \sum_{t= K_{n,j-1} + 1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar X_n)\Big)^2
  \end{align*}
  from the Cauchy-Schwarz inequality. Lemma~\ref{L5}, along
  with~\eqref{eq:17} and~\eqref{eq:25}, implies that
  $\sum_j Z_{nj}^{*2} = O_p(1)$. Lemma~\ref{L3} implies that
  \[
    \tfrac{1}{n} \sum_{j=1}^{J_n} \Big( \sum_{t= K_{n,j-1} + 1}^{K_{nj}}
    (\mu_{nt}^{*} - \bar X_n)\Big)^2 \to^p 0
  \]
  since $\bar X$ itself obeys the \lln.

@@

\begin{proof}
  We will first show that, for every $\epsilon > 0$ and large enough
  $n$, there exists an $\epsilon' > 0$ with the property that
  \begin{equation}\label{eq:31}
    \E \Big(\ind(A) \times \max_{m = 1,\dots,m'} Z^*_n(\tau, m)^2 n / m'\Big) < \epsilon
  \end{equation}
  for any event $A \in \Fs_n \equiv \sigma(X_{n1},\dots,X_{nn})$ with
  $\pr(A) \leq \epsilon'$. Then we will show that this property
  implies uniform integrability.\footnote{%
    The novelty in this part of the proof is the measurability
    requirement. The sigma-field $\Fs_n$ intentionally excludes the
    random variables that define the bootstrap (block length and block
    start periods) and any $A \in \Fs_n$ is independent of those
    variables, so $Z_n^{*}(\tau, m)^2$ is not $\Fs_n$-measurable. If
    $\Fs_n$ were enlarged to include those variables then uniform
    integrability would be an immediate consequence of~\eqref{eq:31}
    and would not need to be shown separately. See \citet[Theorem
    12.9]{Dav:94}.} %
